---
title: "PS71020_Unit1_Homework"
editor: visual
---



::: {.callout-note appearance="simple"}
## Some general notes about the weekly homeworks



*Each week will have an associated homework worksheet in a similar format to this. You should go through the worksheet during your own time to get extra practice. Try to answer the  worksheet questions in your own words however. Model answers will be available within the document itself but please try to answer the questions properly before revealing them. A full R script for the analysis will be published VLE page from Fridays after the labs so you can compare your own script. If your question relates specifically to the lecture content or the homework sheets, please contact Peter Holland by email or use the forum on the VLE page.*
:::

Homework Learning Outcomes
1 - Importing the lecture datasets into R \ 

2 - Perform a basic examination of the dataset \  

3 - Perform basic data wrangling  \ 

4 - Recreate some of the visualisations \  



In this homework worksheet the aim is for you to build a new R script that can perform the above tasks. You can use the code patterns from the lab worksheet and other materials as a guide as well as the questions here.

Here is a summary of the study design:
120 students take part in an 8-week study. Randomly assigned to one of three conditions: \ 

a sleep hygiene programme, \ 

a mindfulness app,  \ 

a control group.  \ 

At the start they log they log their average sleep. 
Every week complete a short wellbeing questionnaire. 
At the end they do a memory test.

There are two .csv files which contain this data, sleep_study.csv contains most of the data, and sleep_study_weekly.csv contains just the weekly wellbeing questionnaire results. Both files can be found on the VLE in the 'Homework Materials' section for this week.

Before downloading the data and starting to write your script what is a good idea to do first? Hint it helps with organisation and keeping materials for one analysis together in one place.

::: {.answer-box}
<input type="text" class="answer-input" placeholder="Type your answer here...">
<button class="reveal-btn" disabled>Show answer</button>

::: {.answer}
Create a new R Project with its own folder to work in. This keeps the data and scripts in the same place to help prevent file location errors.
:::
:::



The first thing we need to do is to load the data. First go to the VLE page and download the two csv files. Now move these to your working directory. 

For desktop RStudio this means moving them to the correct location on your computer. \ 

For RStudio Server this means uploading them to the correct location in your area of the server. \ 


What is the name of the function we have been using to read csv files?

::: {.answer-box}
<input type="text" class="answer-input" placeholder="Type your answer here...">
<button class="reveal-btn" disabled>Show answer</button>

::: {.answer}
read.csv()
:::
:::


```{r}
#| echo: false
sleep_study       <- read.csv("data/sleep_study.csv")
sleep_weekly <- read.csv("data/sleep_study_weekly.csv")
```

In the code used to create this worksheset I have loaded both data sets, I've called the data in sleep_study.csv `sleep_study` and the data in sleep_study_weekly.csv `sleep_weekly`.

If you choose different names that is absolutely fine but remember that any code examples in this worksheet will not match your variable names.

It is good practice for us to examine the data before we try to do any plots.

What R command would I have used to produce the following output?

```{r}
#| echo: false
head(sleep_study)

```
::: {.answer-box}
<input type="text" class="answer-input" placeholder="Type your answer here...">
<button class="reveal-btn" disabled>Show answer</button>

::: {.answer}
head(sleep_study)
:::
:::


The same command produce the following output for the `sleep_weekly` dataframe:

```{r}
#| echo: false
head(sleep_weekly)

```


What is this format referred to when we have one row per subject?

::: {.answer-box}
<input type="text" class="answer-input" placeholder="Type your answer here...">
<button class="reveal-btn" disabled>Show answer</button>

::: {.answer}
Wide format data
:::
:::

We can use the following code to rearrange our data into long format, remembering that we need to use the `library(tidyr)` command to load the **tidyr** package that gives us access to the useful pivot functions.

```{r}
library(tidyr)
sleep_weekly <- pivot_longer(sleep_weekly,
                             cols = week_1:week_8,  # the columns to stack
                             names_to = "week",     # old column names go here...
                             names_prefix = "week_", # ...minus the "week_" part
                             names_transform = as.integer, # so "3" becomes the number 3
                             values_to ="wellbeing")    # the scores go here
dim(sleep_weekly)
head(sleep_weekly)          # 120 students x 8 weeks = 960 rows

```

One important data wrangling step we also need to do is to make sure **R** knows which of our variables is categorical. What are categorical (nominal or ordinal) known as in **R**? What are the categorical variables in our data set?

::: {.answer-box}
<input type="text" class="answer-input" placeholder="Type your answer here...">
<button class="reveal-btn" disabled>Show answer</button>

::: {.answer}
Factors. condition and chronotype
:::
:::

We can use the following code to accomplish this:
```{r}
# condition is a categorical variable (a factor) and fix the order of the groups
#since we will use this more than once we should create it as a variable
cond_levels <- c("Control", "Sleep hygiene", "Mindfulness") 
sleep_study$condition  <- factor(sleep_study$condition,  
                                 levels = cond_levels) # use our newly
sleep_weekly$condition <- factor(sleep_weekly$condition, 
                                 levels = cond_levels)
sleep_study$chronotype <- factor(sleep_study$chronotype)

```

One other thing that can also be good practice is to use the same colours throughout 
all the plots that use the same data groupings. So for example the **control** group will
always appear gray. We can specify this here using the line of code below. Now if we want to re-use these colours we can simply refer to the variable `cond_cols`.


```{r}
# One colour per condition (close to Okabe-Ito colours: safe for colour-blind viewers)
cond_cols <- c("Control" = "darkgray", "Sleep hygiene" = "deepskyblue4", "Mindfulness" = "darkgoldenrod2") 

```

You can look for a guide to a sample of these named colours here <https://r-graph-gallery.com/42-colors-names.html>. However, R contains 657 different named colours, you can print a full list of them to the console by type `colors()`. If you type the name of a colour in "" marks into the console or a script **RStudio** will colour that text with that colour as a quick guide for you.


During the Week 1 Lab Worksheet we used the `hist()` function to plot a histogram of the `mydata$Ladder.score` variable and also saw how to change the colour of the bars and add axis labels and titles. Try to use similar code to reproduce the figure below and add this to your analysis script.

```{r}
#| echo: false
hist(sleep_study$sleep_hours,
     breaks = 12,
     col    = "cornflowerblue",
     main   = "Hours of sleep per night",
     xlab   = "Hours of sleep",
     ylab   = "Number of students")
```
Do you notice something that is not ideal about the y axis in this figure?

::: {.answer-box}
<input type="text" class="answer-input" placeholder="Type your answer here...">
<button class="reveal-btn" disabled>Show answer</button>

::: {.answer}
The bars extend beyond the top of the y axis, which is not good practice. This is a frustrating feature of base R plots. In the full answer code for this worksheet I will provide a solution to this.
:::
:::

If we want to plots the counts of a nominal variable we need to know how many fall into each category. Here we can use a function called **table()** and assign these counts to a variable called `chrono_counts` that we can use for plotting.

```{r}
chrono_counts <- table(sleep_study$chronotype)
chrono_counts

```
The **barplot()** function works very similarly to the `hist()` function we have just used. Try to look at the help for `barplot()` and try to reproduce the figure below. Don't worry about matching the colours exactly, just try to have different colours for the two bars.

```{r}
#| echo: false
barplot(chrono_counts,
        ylim = c(0, max(pretty(chrono_counts))),   # axis reaches above the tallest bar
        col  = c("Evening" = "plum3", "Morning" = "deepskyblue2"), 
        main = "Chronotype",
        xlab = "Chronotype",
        ylab = "Number of students")
```


The next plot we are going to use is a boxplot. In the [Tutorial 5](Pre_sessional_4.qmd) you can see an example of how to use a formula object to create a boxplot. Here the formula we want to use is `wellbeing_final ~ condition`. So we can define it like this:

```{r}
myFormula <- wellbeing_final ~ condition
```

See if you can figure out the code needed to create the following plot:

```{r}
#| echo: false
boxplot(myFormula, data = sleep_study,
        border  = cond_cols,
        ylim    = c(20, 75),
        outline = FALSE,
        main    = "Box plot (median & IQR)",
        xlab    = "Condition",
        ylab    = "Wellbeing at week 8")
```
The final plot we are going to create is the line plot of the wellbeing scores for each week but split by group. We first need to calculate the means for each week and each group, this can be done using the following code:

```{r}
# Calculate the mean wellbeing for each condition in each week
weekly_means <- aggregate(wellbeing ~ condition + week, data = sleep_weekly, FUN = mean)

```

Please don't worry if this looks complex, we will cover how to use this function in future weeks. However, if we look at the output we can see it has done what we would like:

```{r}
head(weekly_means)
```

We now want to get the means for each condition separately. In [Tutorial 5](Pre_sessional_4.qmd) you can see an example of how we can select certain rows of a dataframe, for example only those rows belonging to a certain condition. Here we do this three times to get our three separate sets of weekly means:

```{r}
# Split the means into three separate tables, one per condition
control_means      <- weekly_means[weekly_means$condition == "Control", ]
hygiene_means      <- weekly_means[weekly_means$condition == "Sleep hygiene", ]
mindfulness_means  <- weekly_means[weekly_means$condition == "Mindfulness", ]
```

Now the trick to plotting all three on one graph is to create an empty plot and then draw them one by one. Below is the code to create the plot and draw the first line:

```{r}
# Draw an empty plot with the right axes
plot(weekly_means$week, weekly_means$wellbeing,
     type = "n",
     xlim = c(1, 8),
     ylim = c(40, 56),
     main = "Wellbeing over the 8-week study",
     xlab = "Week",
     ylab = "Mean wellbeing score")

# Add one line per condition
lines(control_means$week, control_means$wellbeing,
      type = "b", pch = 19, lwd = 2, col = cond_cols["Control"])
```

See if you can write the code to draw the other two lines following the same pattern as `lines(control_means$week, control_means$wellbeing, type = "b", pch = 19, lwd = 2, col = cond_cols["Control"])`.

Finally, we can add a legend to the plot by adding the line `legend("topleft", legend = cond_levels, col = cond_cols, lwd = 2, pch = 19, bty = "n")`


The final plot you create should look like this:
```{r}
#| echo: false
# Draw an empty plot with the right axes
plot(weekly_means$week, weekly_means$wellbeing,
     type = "n",
     xlim = c(1, 8),
     ylim = c(40, 56),
     main = "Wellbeing over the 8-week study",
     xlab = "Week",
     ylab = "Mean wellbeing score")

# Add one line per condition
lines(control_means$week, control_means$wellbeing,
      type = "b", pch = 19, lwd = 2, col = cond_cols["Control"])

lines(hygiene_means$week, hygiene_means$wellbeing,
      type = "b", pch = 19, lwd = 2, col = cond_cols["Sleep hygiene"])

lines(mindfulness_means$week, mindfulness_means$wellbeing,
      type = "b", pch = 19, lwd = 2, col = cond_cols["Mindfulness"])

legend("topleft", legend = cond_levels, col = cond_cols, lwd = 2, pch = 19, bty = "n")
```
